---
title: "SV calling from long reads in a single individual"
date: "`r Sys.Date()`"
output: html_document
params:
  wdir: ""
  genome: ""
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, include=FALSE}
require(rmarkdown)
require(vcfR)
require(adegenet)
require(poppr)
require(jsonlite)
require(kableExtra)
require(ggplot2)
require(ggpubr)
require(stringr)
library(tidyverse)
```


```{r discovery, include=FALSE}
# Resolve the report parameters. finalQC.R passes them via render(params = ...),
# but historically the chunks relied on those names leaking in from the caller's
# environment. Prefer params when set, fall back to the inherited globals, so the
# document renders both from finalQC.R and standalone.
if (!is.null(params$wdir) && nzchar(params$wdir)) wdir <- params$wdir
if (!is.null(params$genome) && nzchar(params$genome)) genome <- params$genome

# knitr sets the working directory to this file's location, hence the ../../ prefix.
path_in <- function(...) paste0("../../", wdir, "/", ...)

assembly_path  <- path_in("genome/", genome, "_assembly_data_report.jsonl")
seqreport_path <- path_in("genome/", genome, "_sequence_report.jsonl")
calls_path     <- path_in(genome, "_final.tsv")
unmerged_path  <- path_in("merging_QC/", genome, "_unmerged_sv.tsv")

have_assembly  <- file.exists(assembly_path)
have_seqreport <- file.exists(seqreport_path)
have_unmerged  <- file.exists(unmerged_path)

# The merged callset is a declared input of rule final_report, so its absence is a
# pipeline bug rather than a supported configuration: fail immediately and clearly.
if (!file.exists(calls_path)) {
  stop("Merged callset not found at ", calls_path,
       ". rule final_report declares it as an input; check that rule vcf_to_tsv ran.")
}

# Report the shape of the run rather than assuming it: the set of aligner+caller
# combinations is whatever workflow/scripts/vcf_to_tsv.sh wrote into the header of
# <genome>_final.tsv. Its column layout is 6 fixed fields, FORMAT, one column per
# aligner+caller, then the merged/genotyped sample last.
FORMAT_COL <- 7

merged <- read.table(calls_path, header = TRUE, sep = "\t", check.names = FALSE)

tool_cols <- names(merged)[(FORMAT_COL + 1):(ncol(merged) - 1)]
tools <- data.frame(
  column = tool_cols,
  # Tool columns are named <aligner>_<caller>; aligner names carry no underscore.
  aligner = sub("_.*$", "", tool_cols),
  caller = sub("^[^_]*_", "", tool_cols),
  stringsAsFactors = FALSE
)
aligners <- unique(tools$aligner)
callers <- unique(tools$caller)

fix.merged <- merged[, 1:(FORMAT_COL - 1)]
gt.merged <- merged[, FORMAT_COL:ncol(merged)]

fix.merged$sv_type <- as.vector(gsub("SVTYPE=", "", str_match(fix.merged$INFO, "SVTYPE=[A-Z]+")))
fix.merged$sv_length <- as.numeric(as.vector(gsub("SVLEN=", "", str_match(fix.merged$INFO, "SVLEN=[-]*[0-9]+"))))
fix.merged$supp_vec <- as.vector(gsub("SUPP_VEC=", "", str_match(fix.merged$INFO, "SUPP_VEC=[0-1]+")))
fix.merged$n_supp <- as.numeric(as.vector(gsub("SUPP=", "", str_match(fix.merged$INFO, "SUPP=[0-9]+"))))
fix.merged$jasmine <- ifelse(!is.na(fix.merged$supp_vec), 1, 0)

# Bit i of Jasmine's SUPP_VEC corresponds to tools$column[i]: both orders are set
# by the order the VCFs are listed in workflow/rules/merging.smk. Decode into a
# logical matrix once; every per-tool summary below is derived from it.
supp_raw <- fix.merged$supp_vec
supp_ok <- !is.na(supp_raw) & nchar(supp_raw) == nrow(tools)
if (any(!is.na(supp_raw) & nchar(supp_raw) != nrow(tools))) {
  warning(
    "SUPP_VEC length does not match the ", nrow(tools),
    " tool columns of ", basename(calls_path),
    "; those calls are treated as unsupported."
  )
}
supp_mat <- matrix(FALSE,
  nrow = length(supp_raw), ncol = nrow(tools),
  dimnames = list(NULL, tools$column)
)
if (any(supp_ok)) {
  supp_mat[supp_ok, ] <- do.call(rbind, strsplit(supp_raw[supp_ok], "")) == "1"
}

# Facet layouts follow the number of callers instead of a fixed 4 columns.
facet_ncol <- max(1, length(callers))
```


# Assembly


## Assembly summary statistics

```{r assemblysummary, echo = F}
if (!have_assembly) {
  cat("No assembly report found at", assembly_path, "- section skipped.\n")
} else {
json = jsonlite::read_json(assembly_path)

cat("Organism name:", json$organism$organismName, "\n")

cat("Assembly name:", json$assemblyInfo$assemblyName, "\n")
cat("Accession:", json$accession, "\n")

cat("Bioproject Accession:", json$assemblyInfo$bioprojectAccession, "\n")

cat("Total sequence length:", json$assemblyStats$totalSequenceLength, "\n")
cat("Total ungapped length:", json$assemblyStats$totalUngappedLength, "\n")

cat("Genome coverage:", json$assemblyStats$genomeCoverage, "\n")

cat("Number of contigs:", json$assemblyStats$numberOfContigs, "\n")
cat("Contig N50 length (Mb):", json$assemblyStats$contigN50/10^6, "\n")

cat("Number of scaffolds:", json$assemblyStats$numberOfScaffolds, "\n")
cat("Scaffold N50 length (Mb):", json$assemblyStats$scaffoldN50/10^6, "\n")

cat("GC percent:", json$assemblyStats$gcPercent, "\n")
}
```

## Chromosomes


```{r chromosomes, echo = F}
# genome_size stays NA when the sequence report is absent; the plots below drop the
# genome-relative axis rather than failing.
genome_size = NA
chr = NULL

if (!have_seqreport) {
  cat("No sequence report found at", seqreport_path, "- section skipped.\n")
} else {
dat = tidyjson::read_json(seqreport_path, format = "jsonl")
chr = data.frame(idx = dat$document.id)

chr$seqname = NA
chr$genbankaccession = NA
chr$type = NA
chr$length = NA
chr$gcpercent = NA

for (i in 1:nrow(chr)) {
  chr$seqname[i] = dat$..JSON[[i]]$chrName
  chr$length[i] = dat$..JSON[[i]]$length/10^6
  chr$type[i] = dat$..JSON[[i]]$assignedMoleculeLocationType
  chr$gcpercent[i] = ifelse(!is.null(dat$..JSON[[i]]$gcPercent), dat$..JSON[[i]]$gcPercent, NA)
  chr$genbankaccession[i] = dat$..JSON[[i]]$genbankAccession
}

genome_size = sum(chr$length)
}
```



```{r chromosometable, echo=FALSE, message=FALSE, warning=FALSE}
if (!is.null(chr)) {
kable(chr[,-1], digits = 2, align = "c", caption = "Summary of chromosome assembly.", col.names = c("Name", "Genbank Accession", "Type", "Length (Mb)", "GC percent"), format = "html", table.attr = "style='width:80%;'") %>% kableExtra::kable_styling()
}
```


# Mapping


```{r mapping, echo = F}
# One block per aligner actually present, so a run with a single aligner reports
# that aligner instead of failing on the missing one.
if (!length(aligners)) {
  cat("No mapping statistics found in", path_in("mapping_QC"), "\n")
}

for (aligner in aligners) {
  stats_path = path_in("mapping_QC/", genome, "_", aligner, "_mapping.stats.tsv")

  if (!file.exists(stats_path)) {
    cat("No mapping statistics for", aligner, "- skipped.\n")
    next
  }

  print(paste("Mapping with", aligner))
  print(gsub("\t", " ", readLines(stats_path)))
}
```




The diagnostic plots for the mapping are in the `mapping_QC/` directory, one set per aligner.



# SV calling


The final VCF produced contains one sample per aligner + SV caller pair, plus one more:


* each of those samples is the individual callset of a pair of aligner + SV caller
* the last sample is the final genotype call after merging callsets with Jasmine


Below are presented the callsets for the different tools (aligner  + SV caller) before the merging with Jasmine. Hence some calls may not be represented in the final sample (lost during merging).


The callset of each individual tool has been filtered and genotyped with SVjedi-graph, yielding a Phred-scale log likelihood and a consistent quality score. During filtering, SV were filtered by maximum SV length (default = 30Mb), minimum allele depth (default = 5), min depth of the variant (default = 10) and max depth of the variant (default = 500), and max phred-scale likelihood PL (default = 60).




```{r get_tool_support, echo = FALSE}
# Per-aligner and per-caller support, collapsed from the SUPP_VEC matrix built in
# the discovery chunk. A call counts as supported by an aligner (or a caller) when
# any tool using it supports the call.
for (aligner in aligners) {
  fix.merged[[aligner]] <- as.integer(
    rowSums(supp_mat[, tools$aligner == aligner, drop = FALSE]) > 0
  )
}

for (caller in callers) {
  fix.merged[[caller]] <- as.integer(
    rowSums(supp_mat[, tools$caller == caller, drop = FALSE]) > 0
  )
}
```



### Test if merging all samples worked

After merging, the merged callset is merged with all individual tools callset. This test checks if bcftools has been able to merge correctly individual and merged callsets. The individual callsets and merged callsets should be merged even if calls are genotyped as missing "./.".


```{r testMerging, echo=FALSE, results=FALSE, message=FALSE, warning=FALSE}
# Rebuild the support vector from the genotype columns and compare it with the one
# Jasmine recorded. Both are assembled in tools$column order, whatever that order is.
pattern = "./.:NA:NA:NA:NA:.:.:."

observed = vapply(
  tools$column,
  function(tool) ifelse(grepl(pattern, merged[[tool]], fixed = TRUE), "0", "1"),
  character(nrow(merged))
)
observed = matrix(observed, nrow = nrow(merged), dimnames = list(NULL, tools$column))

valid_supp_vec = apply(observed, 1, paste0, collapse = "")
```

```{r, echo = T}
cat("The percentage of calls successfully merged:", sum(valid_supp_vec == fix.merged$supp_vec, na.rm = TRUE) / nrow(fix.merged) * 100, "%\n")
```



### Parsing the merged callset

```{r parseSV, echo = F}
# Prefer OLDTYPE (the pre-merging type Jasmine preserves) but keep the SVTYPE parsed
# in the discovery chunk where OLDTYPE is absent, rather than blanking every type.
old_type = as.vector(gsub("OLDTYPE=", "", str_match(fix.merged$INFO, "OLDTYPE=[A-Z]+")))
fix.merged$sv_type = ifelse(is.na(old_type), fix.merged$sv_type, old_type)
fix.merged$sv_length = as.numeric(as.vector(gsub("SVLEN=", "", str_match(fix.merged$INFO, "SVLEN=[+-]*[0-9]+"))))
```


```{r, echo = F}
# One block per tool, selected on that tool's own SUPP_VEC bit.
#
# NOTE this is a deliberate change of behaviour. The previous version selected
# `aligner == 1 & caller == 1`, i.e. "supported by some tool using this aligner AND
# by some tool using this caller", which also picked up calls the specific pair never
# made (e.g. a call from minimap2_svim and ngmlr_sniffles counted towards
# minimap2_sniffles). Per-tool counts in the figures below therefore change.
tidy_fix = do.call(rbind, lapply(seq_len(nrow(tools)), function(i) {
  rows = fix.merged[supp_mat[, tools$column[i]], , drop = FALSE]
  if (!nrow(rows)) {
    return(NULL)
  }
  rows$aligner = tools$aligner[i]
  rows$caller = tools$caller[i]
  rows
}))
```



## Number of SV detected by each caller

```{r SVcalling, echo = T}
table(tidy_fix$aligner)
table(tidy_fix$caller)

table(tidy_fix$aligner, tidy_fix$caller)
```



## SV quality score


*Note that all SV callers do not use the same way of calculating the Quality Score and they are not all on the same scale (e.g. log/phred-scale).* Consequently, the QUAL score plotted here is the SVjedi-graph genotyping QUAL score, estimated in a consistent way across SV callers.



```{r SVqualityAligner, echo = F, warning = F, fig.cap = "SVjedi-graph quality score for each aligner."}
ggplot(tidy_fix, aes(x = aligner, y = as.numeric(QUAL), colour = aligner, group = aligner)) +
  geom_jitter(alpha = 0.1) +
  geom_boxplot(alpha = 0.1, width = 0.1, fill = NA, color = "black") +
  geom_violin(width = 1.5, fill = NA, color = "black") +
  coord_trans(y = "log10") +
  xlab("Aligner") + ylab("Quality Score") +
  theme_bw()
```



```{r SVqualityCaller, echo = F, warning = F, fig.cap = "SVjedi-graph quality score for each aligner and SV caller."}
ggplot(tidy_fix, aes(x = caller, y = as.numeric(QUAL), colour = caller, group = caller)) +
  geom_jitter(alpha = 0.1) +
  geom_boxplot(alpha = 0.1, width = 0.1, fill = NA, color = "black") +
  geom_violin(width = 1.5, fill = NA, color = "black") +
  coord_trans(y = "log10") +
  xlab("SV caller") + ylab("Quality Score") +
  facet_wrap(~ aligner, scales = "fixed", nrow = max(1, length(aligners))) +
  theme_bw()
```


```{r SVqualitydensity, echo = F, warning = F, fig.cap = "Density distribution of SVjedi-graph quality scores for each aligner and SV caller."}
ggplot(tidy_fix, aes(x = as.numeric(QUAL), colour = caller, group = caller)) +
  geom_density() +
  xlab("Quality Score") + ylab("Density") +
  coord_trans(x = "log10") +
  facet_wrap(~ aligner + caller, scales = "free", ncol = facet_ncol) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```



## SV type

```{r SVtypetable, echo = T}
table(tidy_fix$sv_type, tidy_fix$caller)
```



```{r SVtype, echo = F, warning = F, fig.cap = "Number of SVs of each type for each aligner and SV caller."}
ggplot(tidy_fix, aes(x = sv_type, fill = sv_type, colour = sv_type)) +
  geom_bar() +
  xlab("SV type") + ylab("Count") + labs(fill = "SV type", colour = "SV type") +
  facet_wrap(~ aligner + caller, scales = "fixed", ncol = facet_ncol) +
  theme_bw()
```


## SV length


```{r SVlength, echo = F, warning = F, fig.cap="Total absolute length (left axis) and proportion of the genome (right axis) covered by SVs. The black line is the total genome size."}
df = aggregate(sv_length ~ sv_type + caller + aligner, tidy_fix, function(x) {sum(abs(x))})

df_sample_size = aggregate(sv_length ~ sv_type + caller + aligner, tidy_fix, function(x) {length(x)})
df$sample_size = df_sample_size$sv_length
  
df$proportion_genome = df$sv_length/(genome_size * 10^6)

p = ggplot(df, aes(x = sv_type, y = sv_length/10^6, fill = sv_type, colour = sv_type)) +
  geom_bar(stat = "identity") +
  xlab("SV type") + ylab("Total absolute length (Mb)") +
  labs(fill = "SV type", colour = "SV type") +
  facet_wrap(~ aligner + caller, scales = "fixed", ncol = facet_ncol) +
  theme_bw()

# The genome-relative axis needs the sequence report; drop it when unavailable
# rather than failing on a NA scale.
if (!is.na(genome_size)) {
  p = p +
    geom_hline(yintercept = genome_size) +
    scale_y_continuous(
      name = "Total absolute length (Mb)",
      sec.axis = sec_axis(trans = ~ . / genome_size, name = "Proportion of the genome")
    )
}

p
```






# Merging with Jasmine


How many calls were merged with jasmine?


```{r, echo = TRUE}
cat("Absolute number of calls after merging with Jasmine:", sum(fix.merged$jasmine), "\n")

cat("Proportion of calls after merging with Jasmine:", sum(fix.merged$jasmine) / nrow(fix.merged), "\n")
```


Among these calls, how many SVs remained unmerged?

We consider that variants overlapping with less than 25 bp of difference in start/end positions are unmerged variants.
It can be variants of different types (e.g. DEL vs INS).

```{r, echo = TRUE}
if (!have_unmerged) {
  cat("No unmerged-SV report found at", unmerged_path, "- section skipped.\n")
} else {
unmerged = read_tsv(unmerged_path)

cat("Absolute number of calls UNMERGED:", nrow(unmerged), "\n")

cat("Proportion of calls UNMERGED:", nrow(unmerged) / nrow(fix.merged), "\n")
}
```



## Agreement between tools


How many tools support each call, and which combinations agree. This replaces the
previous Euler diagrams, which could only be drawn for a fixed set of four callers.


```{r cosupport, echo = F}
cat("Number of calls by count of supporting tools:\n")
print(table(n_tools = rowSums(supp_mat)))
```


```{r cosupport_by_type, echo = F}
cat("\nCalls supported by each tool, per SV type:\n")
# Share one level set across tools, so the result is a rectangular table even when
# a tool never calls a given SV type.
sv_levels = sort(unique(na.omit(fix.merged$sv_type)))
print(
  sapply(tools$column, function(tool) {
    table(factor(fix.merged$sv_type[supp_mat[, tool]], levels = sv_levels))
  })
)
```


```{r cosupport_aligner_caller, echo = F}
cat("\nCalls supported by each aligner and caller:\n")
print(table(tidy_fix$aligner, tidy_fix$caller))
```




# Genotyping and final callset


After merging the callsets and re-genotyping, all SV called are represented in the final callset. They have not been filtered. It is left to the user to filter them afterward, depending on the question.



```{r samplename, echo = F}
colnames(gt.merged)[-1]
```



```{r genotyping, echo = F}
# Only the final callset, all tools merged: FORMAT plus the last sample, whatever
# the number of tool columns in between.
gt.geno = gt.merged[, c(1, ncol(gt.merged))]
```


## SV type and length


```{r SVtablefinal}
table(fix.merged$sv_type)
```


The stacked barplot of SV count for each SV type and SV length (no filter on quality scores and quality score > 10).

```{r SVtypefinal, echo = F, warning = F, fig.cap = "Joint distribution of SV length and number of SV, per SV type."}
p1 = ggplot(fix.merged, aes(x = abs(sv_length), fill = sv_type, colour = sv_type)) +
  geom_bar() +
  xlab("Length of Structural Variant (bp), less than 1 kb") + ylab("SV Count") +
  xlim(0, 1000) +
  labs(fill = "SV type", colour = "SV type") +
  theme_bw()

p2 = ggplot(fix.merged, aes(x = abs(sv_length), fill = sv_type, colour = sv_type)) +
  geom_bar() +
  xlab("Length of Structural Variant (bp), less than 50 kb") + ylab("SV Count") +
  xlim(0, 50000) +
  labs(fill = "SV type", colour = "SV type") +
  theme_bw()


p3 = ggplot(fix.merged, aes(x = abs(sv_length), fill = sv_type, colour = sv_type)) +
  geom_bar() +
  xlab("Length of Structural Variant (bp), unlimited scale") + ylab("SV Count") +
  labs(fill = "SV type", colour = "SV type") +
  theme_bw()



ggarrange(p1, p2, p3, nrow = 3)
```



```{r SVlengthfinal, echo = F, warning = F, fig.cap = "SV length as a function of SV type."}
fix.merged$sv_type = as.factor(fix.merged$sv_type)

p1 = ggplot(fix.merged, aes(y = abs(sv_length) / 1000, x = sv_type, fill = sv_type)) +
  geom_jitter(aes(color = sv_type), alpha = 0.1) +
  geom_boxplot(alpha = 0.5) +
  ylab("SV Length (kb)") + xlab("SV type") +
  labs(fill = "SV type", colour = "SV type") +
  theme_bw()


p2 = ggplot(fix.merged, aes(y = log10(abs(sv_length)), x = sv_type, fill = sv_type)) +
  geom_jitter(aes(color = sv_type), alpha = 0.1) +
  geom_boxplot(alpha = 0.5) +
  ylab("SV Length (log10 scale)") + xlab("SV type") +
  labs(fill = "SV type", colour = "SV type") +
  theme_bw()

ggarrange(p1, p2, nrow = 2)
```



## Missing data


```{r parsinggeno, echo=FALSE, message=FALSE, warning=FALSE}
# Look the subfields up by name in the FORMAT column rather than at fixed offsets,
# so the report survives a genotyper emitting a different FORMAT layout.
format_keys = strsplit(as.character(gt.geno[[1]][1]), ":")[[1]]

get_field = function(values, key) {
  i = match(key, format_keys)
  if (is.na(i)) {
    warning("FORMAT field '", key, "' is absent; related figures will be empty.")
    return(rep(NA_character_, length(values)))
  }
  vapply(
    strsplit(as.character(values), ":"),
    function(v) if (length(v) >= i) v[i] else NA_character_,
    character(1)
  )
}

# Split a comma-separated subfield (AD, PL) into its nth value.
get_part = function(values, n) {
  as.numeric(vapply(
    strsplit(values, ","),
    function(v) if (length(v) >= n) v[n] else NA_character_,
    character(1)
  ))
}

gt.geno$GT = get_field(gt.geno[[2]], "GT")
gt.geno$AD = get_field(gt.geno[[2]], "AD")
gt.geno$DP = get_field(gt.geno[[2]], "DP")
gt.geno$PL = get_field(gt.geno[[2]], "PL")

gt.geno$AD_REF = get_part(gt.geno$AD, 1)
gt.geno$AD_ALT = get_part(gt.geno$AD, 2)
gt.geno$PL_HOMREF = get_part(gt.geno$PL, 1)
gt.geno$PL_HET = get_part(gt.geno$PL, 2)
gt.geno$PL_HOMALT = get_part(gt.geno$PL, 3)
```


```{r missinggeno}
cat("Proportion of missing data (./.):", sum(gt.geno$GT == "./.")/nrow(gt.geno), "\n")

cat("Proportion of partially missing data (./?):", sum(gt.geno$GT %in% c("./0", "0/.", "1/.", "./1"))/nrow(gt.geno), "(should be 0 if everything is ok) \n")
```



## Heterozygosity


```{r HeGeno}
cat("Proportion of heterozygous SV (0/1 or 1/0):", sum(gt.geno$GT %in% c("1/0", "0/1"))/nrow(gt.geno), "\n")
```


We investigated the usual Variant Calling quality criterion:

* DP: Approximate read depth; some reads may have been filtered
* AD: Allelic depths for the ref and alt alleles in the order listed
* PL: Normalized, Phred-scaled likelihoods for genotypes as defined in the VCF specification (lower score is better, lower score = most likely genotype)


## Supporting reads (DP:AD)



Approximate read depth (should be balanced between reference and variant if unbiased).


```{r SV_DP, echo = F, warning = F, fig.cap = "Genotype quality. Variant allele read depth (AD) as a function of the genotype."}
ggplot(gt.geno, aes(x = AD_REF, y = AD_ALT, group = GT, colour = GT)) +
  geom_point(alpha = 0.3) +
  xlab("Reference allele read depth (AD)") + ylab("Variant allele read depth (AD)") +
  theme_bw()
```



```{r SV_DP_500, echo = F, warning = F, fig.cap = "Genotype quality. Variant allele read depth (AD) as a function of the genotype. Subset of read depth < 500 for details."}
ggplot(gt.geno, aes(x = AD_REF, y = AD_ALT, group = GT, colour = GT)) +
  geom_point(alpha = 0.3) +
  xlab("Reference allele read depth (AD) (< 500)") + ylab("Variant allele read depth (AD) (< 500)") +
  xlim(0, 500) + ylim(0, 500) +
  theme_bw()
```


Variant allelic depth of the ALT allele (higher for heterozygotes is better).


```{r AD_GT, echo = F, warning = F}
gt.geno$GT = as.factor(gt.geno$GT)

p1 = ggplot(gt.geno, aes(x = GT, y = AD_ALT, group = GT, colour = GT)) +
  geom_jitter(aes(color = GT), alpha = 0.1) +
  geom_boxplot(alpha = 0.5, color = "black") +
  ylab("Variant allelic depth (AD)") +
  theme_bw()

p2 = ggplot(gt.geno, aes(x = GT, y = AD_ALT, group = GT, colour = GT)) +
  geom_jitter(aes(color = GT), alpha = 0.1) +
  geom_boxplot(alpha = 0.5, color = "black") +
  ylab("Variant allelic depth (AD)") +
  ylim(0, 100) +
  theme_bw()

ggarrange(p1, p2, ncol = 2)
```


## Phred-scale likelihoods for heterozygotes



Scores for 0/0 (homozygous ref), 0/1 (heterozygous), and 1/1 (homozygous alt) genotypes. For a phred-scaled likelihood of P, the raw likelihood of that genotype L = 10-P/10 (so the higher the number, the less likely it is that your sample is that genotype). A lower phred-scale likelihood is better. The sum of likelihoods is not necessarily 1. 



```{r PL, echo = F, warning = F}
df = data.frame(GT = c(rep("0/0", sum(gt.geno$GT %in% c("0/0"))),
                       rep("0/1", sum(gt.geno$GT %in% c("0/1"))),
                       rep("1/1", sum(gt.geno$GT %in% c("1/1")))),
                PL = c(gt.geno$PL_HOMREF[which(gt.geno$GT %in% c("0/0"))],
                       gt.geno$PL_HET[which(gt.geno$GT %in% c("0/1"))],
                       gt.geno$PL_HOMALT[which(gt.geno$GT %in% c("1/1"))]))
df$GT = as.factor(df$GT)


p1 = ggplot(df, aes(x = GT, y = PL, group = GT, colour = GT)) +
  geom_jitter(aes(color = GT), alpha = 0.1) +
  geom_boxplot(alpha = 0.5, color = "black") +
  ylab("Phred-scaled genotype likelihood") +
  theme_bw()

p2 = ggplot(df, aes(x = GT, y = PL, group = GT, colour = GT)) +
  geom_jitter(aes(color = GT), alpha = 0.1) +
  geom_boxplot(alpha = 0.5, color = "black") +
  ylab("Phred-scaled genotype likelihood") +
  ylim(0, 100) +
  theme_bw()

ggarrange(p1, p2, ncol = 2)
```


Now we plot the likelihood of the heterozygote for the three genotypes inferred. We expect a very low PL value for hetozygotes while PL values should be higher for HOM genotypes that are most likely calling errors. The difference in PL values between HET and HOM genotypes can be leveraged as a cutoff to filter low quality SV calls.



```{r PLHET, echo = F, warning = F}
df = data.frame(GT = c(rep("0/0", sum(gt.geno$GT %in% c("0/0"))),
                       rep("0/1", sum(gt.geno$GT %in% c("0/1"))),
                       rep("1/1", sum(gt.geno$GT %in% c("1/1")))),
                PL = c(gt.geno$PL_HET[which(gt.geno$GT %in% c("0/0"))],
                       gt.geno$PL_HET[which(gt.geno$GT %in% c("0/1"))],
                       gt.geno$PL_HET[which(gt.geno$GT %in% c("1/1"))]))
df$GT = as.factor(df$GT)


p1 = ggplot(df, aes(x = GT, y = PL, group = GT, colour = GT)) +
  geom_jitter(aes(color = GT), alpha = 0.1) +
  geom_boxplot(alpha = 0.5, color = "black") +
  ylab("HET Phred-scaled genotype likelihood") +
  theme_bw()

p2 = ggplot(df, aes(x = GT, y = PL, group = GT, colour = GT)) +
  geom_jitter(aes(color = GT), alpha = 0.1) +
  geom_boxplot(alpha = 0.5, color = "black") +
  ylab("HET Phred-scaled genotype likelihood") +
  ylim(0, 100) +
  theme_bw()

ggarrange(p1, p2, ncol = 2)
```





# Precision/recall for each tool and each type of SV


Estimate the calling rates for each tool:


* the TRUE POSITIVE is when a call is genotyped as HET by SVjedi-graph
* the FALSE POSITIVE is when a call is genotyped as HOM by SVjedi-graph
* the FALSE NEGATIVE is when a call is genotyped as missing (./.) while it has been called as HET for at least one other tool (we assume that a HET call is a true call, yet it an approximation and ground truth remains unknown). Note that because of this approximation we might be underestimating the false negative rate (the proportion of undetected variants) and overestimating the recall and F1 score. However, without ground truth, it is difficult to estimate the false negative rate, and our performance score is still valid to compare and weight the different tools.
* the TRUE NEGATIVE is not estimated, and is not used in precision/recall calculation

Quick reminder, the precision is TP / (TP + FP) and recall is TP / (TP +  FN). The F1 score is then F1 = 2 * ((precision * recall) / (precision + recall)).


IMPORTANT NOTE. Since the ground truth is unknown, TP, FP and FN are approximations in this context, making some assumptions:


* the empirical ground truth depends on our filtering criteria and the definition of a TP (the TP is overestimated if some HET calls are in fact false positives, i.e. if the genotyping error rate is > 0, which is most likely). You can set a more stringent filtering to reduce this bias.
* we assume that all true SVs in this individual has been called by at least one SV caller (the FN is underestimated if SVs have not been called by at least one tool)



We are trying to estimate the true precision/recall of each tool. We want to be able to estimate the relative performance of each tool compared to the others, for a given dataset.
These empirical performance scores are useful to give more importance to some tools and to have a proxy of the confidence we can have in a callset.
They can be used downstream to weigth any measure of confidence you have in a call (e.g. a genotype likelihood), depending on which tools called it.


```{r, echo = T}
# Build a matrix of calls per tool, one column per entry in `tools`.
genotype = fix.merged[,c(1:3, 7)]

call_class = function(values) {
  gen = vapply(strsplit(as.character(values), ":"), `[`, character(1), 1)
  ifelse(gen == "0/1", "HET", ifelse(gen == "./.", "MISSING", "HOM"))
}

for (tool in tools$column) {
  genotype[[tool]] = call_class(gt.merged[[tool]])
}

# Estimate precision, recall and F1 score
performance = data.frame(tool = tools$column,
                         n_calls = NA,
                         true_positive = NA,
                         false_positive = NA,
                         false_negative = NA,
                         stringsAsFactors = FALSE)

n_calls = nrow(genotype)

het_in_one_tool = apply(genotype[, tools$column, drop = FALSE], 1, function(x) {"HET" %in% x})

for (i in 1:nrow(performance)) {
  performance$n_calls[i] = sum(genotype[,performance$tool[i]] != "MISSING")
  performance$true_positive[i] = sum(genotype[,performance$tool[i]] == "HET")
  performance$false_positive[i] = sum(genotype[,performance$tool[i]] == "HOM")
  performance$false_negative[i] = sum(genotype[,performance$tool[i]] == "MISSING" & het_in_one_tool)
}

performance$precision = performance$true_positive / (performance$true_positive + performance$false_positive)
performance$recall = performance$true_positive / (performance$true_positive + performance$false_negative)

performance$f1 = 2 * ((performance$precision * performance$recall) / (performance$precision + performance$recall))
```


<!-- Now estimate the performance for each type of SV (DEL, INS, DUP, INV). -->


```{r, echo = F}
performance$sv_type = "all"


# SV types come from the data rather than a fixed list, so a callset without INV
# (or with a type this pipeline has not seen before) is reported correctly.
for (type in sort(unique(na.omit(genotype$sv_type)))) {
  # Estimate precision, recall and F1 score
  performance_per_type = data.frame(tool = tools$column,
                           n_calls = NA,
                           true_positive = NA,
                           false_positive = NA,
                           false_negative = NA,
                           stringsAsFactors = FALSE)

  genotype_subset = genotype[which(genotype$sv_type == type),]
  # n_calls = nrow(genotype_subset)

  het_in_one_tool = apply(genotype_subset[, tools$column, drop = FALSE], 1, function(x) {"HET" %in% x})

  for (i in 1:nrow(performance_per_type)) {
    performance_per_type$n_calls[i] = sum(genotype_subset[,performance_per_type$tool[i]] != "MISSING")
    performance_per_type$true_positive[i] = sum(genotype_subset[,performance_per_type$tool[i]] == "HET")
    performance_per_type$false_positive[i] = sum(genotype_subset[,performance_per_type$tool[i]] == "HOM")
    performance_per_type$false_negative[i] = sum(genotype_subset[,performance_per_type$tool[i]] == "MISSING" & het_in_one_tool)
  }
  
  performance_per_type$precision = performance_per_type$true_positive / (performance_per_type$true_positive + performance_per_type$false_positive)
  performance_per_type$recall = performance_per_type$true_positive / (performance_per_type$true_positive + performance_per_type$false_negative)
  
  performance_per_type$f1 = 2 * ((performance_per_type$precision * performance_per_type$recall) / (performance_per_type$precision + performance_per_type$recall))
  
  performance_per_type$sv_type = type
  
  performance = rbind(performance, performance_per_type)
}
```



```{r, echo = T, tab.cap = "Performance of the different SV calling tools for each SV type and all SV pooled."}
kable(performance, digits = 3, align = "c")
```





```{r, echo = F}
# Save the performance output. This file is a declared output of rule final_report,
# so it must be written even when the directory does not exist yet (standalone runs;
# under Snakemake the parent directory is created for us).
performance_path = path_in("performance/", genome, "_performance.tsv")
dir.create(dirname(performance_path), showWarnings = FALSE, recursive = TRUE)

write.table(performance, performance_path,
            quote = F, sep = "\t", row.names = F, col.names = T)
```




<!-- 
## Likelihood Ratio Test between the best likelihood and the HET likelihood


An alternative measure of confidence for each call. -->





# SV and genomic landscapes



## Quality and support along the genome

Quality scores and allele support are expected to homogeneous along the genome.


Quality Score.

```{r QClandscape, echo = F, warning = F}
ggplot(fix.merged, aes(x = as.numeric(POS)/10^6, y = as.numeric(QUAL))) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "gam", se = T) +
  xlab("Position (Mb)") + ylab("SV Quality score") +
  facet_wrap(~ CHROM, scales = "free_y", ncol = 3) +
  theme_bw()
```


Support for variant allele.


```{r suppportlandscape, echo = F, warning = F}
fix.merged$sv_support = unlist(lapply(fix.merged$supp_vec, function(x) sum(as.numeric(strsplit(x, "")[[1]]))))

ggplot(fix.merged, aes(x = as.numeric(POS)/10^6, y = as.numeric(sv_support))) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "gam", se = T) +
  xlab("Position (Mb)") + ylab("SV support for allele variant") +
  facet_wrap(~ CHROM, scales = "free_y", ncol = 3) +
  scale_y_continuous(breaks = seq_len(nrow(tools)), labels = as.character(seq_len(nrow(tools)))) +
  theme_bw()
```






## The distribution of SVs along the genome


```{r SVlandscape, echo = F, warning = F}
ggplot(fix.merged, aes(x = as.numeric(POS)/10^6)) +
  geom_histogram(bins = 1000) +
  xlab("Position (Mb)") + ylab("SV count") +
  facet_wrap(~ CHROM, ncol = 3)
```




